---
title: "Single-cell Mapper"
author:
- name: Gennaro Gambardella
  affiliation: TIGEM (Telethon Institute of Genetics and Medicine) 
package: gficf
output:
  BiocStyle::html_document
vignette: |
  %\VignetteIndexEntry{Single-cell Mapper}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Google tag (gtag.js) -->
<script async src="https://www.googletagmanager.com/gtag/js?id=G-XD8QKTWJ1D"></script>
<script>
  window.dataLayer = window.dataLayer || [];
  function gtag(){dataLayer.push(arguments);}
  gtag('js', new Date());

  gtag('config', 'G-XD8QKTWJ1D');
</script>

# Introduction {.unnumbered}

Single-cell Mapper (scMAP) is a transfer learning algorithm that combines text mining data transformation and a k-nearest neighbours’ (KNN) classifier to map a query set of single-cell transcriptional profiles on top of a reference atlas. scMAP consists of three main steps:  

1. In the first step query cell profiles are normalized with the GF-ICF  method but using the ICF weights learned by using the reference atlas;  

2.  In the second step,  normalized cell profiles are first projected into the NMF (or PC)  sub-space of the reference atlas and then mapped into its UMAP embedding space;  

3. Finally, the KNN algorithm is used to contextualize mapped cells and annotate them.  

Please read the manuscript for additional details and estimated performances of the method.

# Building the reference atlas
For this tutorial, we will be using as a reference atlas the `small_BC_atlas dataset` included in the `gficf` package (Figure \@ref(fig:atlas)). This dataset is a small version (comprising only 4,760 cells) of the Breast cancer cell-line atlas we recently published ([Gambardella et al.](https://www.nature.com/articles/s41467-022-29358-6)). 

```{r atlas, fig.cap="UMAP of referece BC Atlas. UMAP plot of the small_BC_atlas dataset where cells are color-coded according to their cell-line of origin.",echo=TRUE,cache=TRUE,warning=FALSE,message=F,prompt=FALSE,results='hide'}
require(gficf)
require(ggplot2)

# Step 1. Build the reference atlas
# Load the RAW UMI count matrix on which to build the reference atlas
data("small_BC_atlas")

# 1.1. Normalization and gene filtering
data <- gficf( M = small_BC_atlas,
               cell_count_cutoff = 15,
               nonz_mean_cutoff = 1.12,
               cell_percentage_cutoff2 = 0.05,
               normalize = T,
               verbose = T)

# 1.2 Create NMF (or PCA) subspace 
# using all genes (usually improves performances)
data <- runNMF(data = data,dim = 50)

# 1.3 Create t-UMAP space
data <-runReduction(data = data,reduction = "umap",nt = 2,verbose = T)

# 1.4 Let's add info about the cell-line of origin
# Cell meta-data can stored in the data$embedded data.frame
# Let' add the info about the cell-line, stripping this information
# from the name of the cell and storing it into ccl column.
data$embedded$ccl = sapply(
                            strsplit(x = rownames(data$embedded),
                                     split = "_",fixed = T)
                            ,function(x) x[1]
                          )

# Plot cells by cell line of origin
p = plotCells(data = data,colorBy = "ccl",pointShape = 19) + 
    xlab("UMAP 1") + 
    ylab("UMAP 2")

plot(p)
```

# How to map new cells into a reference atlas

After we have built our reference atlas (Figure \@ref(fig:atlas)), we use the 930 cells available into the `test_BC_atlas` dataset to test our mapping algorithm (Figure \@ref(fig:map)). Mapping of new cells into the embedded space of the reference atlas is performed by the `scMAP()` function. Metadata and coordinate of new mapped cells are stored into `data$embedded.predicted` data.frame.

```{r map, fig.cap="UMAP of referece BC Atlas and mapped cells. UMAP plot of the small_BC_atlas dataset where cells are color-coded according to their cell-line of origin. Black points are the 930 mapped cells contained in the test_BC_atlas dataset",echo=TRUE,cache=TRUE,warning=FALSE,message=F,prompt=FALSE,results='hide'}

# Step 2. Map new cells into reference atlas.
# 2.1 Load the new cells to map
data("test_BC_atlas")

# 2.2 Cell mapping
data = scMAP(data = data, x = test_BC_atlas,nt = 2,normalize = T,verbose = T)

# 2.3 Plot mapped cells
p = ggplot() + 
    geom_point(data = data$embedded,aes(x=X,y=Y,color=ccl),
               shape=19,size=.5) + 
    geom_point(data = data$embedded.predicted,aes(x=X,y=Y),
               color="black",shape=19,size=.2) + 
    theme_bw() +
    xlab("UMAP 1") + 
    ylab("UMAP 2") +
    theme(legend.position = "none")

plot(p)
```

# How to contextualize mapped cells

Finally we can use KNN to classify and thus contextualize mapped cells into the reference atlas. In this example we try to infer cell-line of origin of each mapped cell for winch. After the mapping we also compute the classification accuracy that is for this example of 93.23%. Specifically, mapping task can be performed with the function `classify.cells()` of `gficf` package. In this function the class of reference cells must be specified by the `classes` parameter. Here we use as class of each cell its cell-line of origin. Classification results are stored into `data$embedded.predicted` data.frame where the column `predicted.class` reports the predicted class and the column `class.prob` its predicted probability.   

The `classify.cells()` function is built on top of [`KernelKnn`](https://github.com/mlampros/KernelKnn) package and thus implement several measures to compute the distances among cells that can be specified trough the `knn_method` parameter. At the same time the parameter `knn_weights_fun` can be used to specify the kernel function to use for cell classification. Default is `knn_weights_fun = NULL` that correspond the to [unweighted KNN algorithm](https://en.wikipedia.org/wiki/K-nearest_neighbors_algorithm). For further details on possible kernel functions to use, plese see this [blog-post](http://mlampros.github.io/2016/07/10/KernelKnn/) of original `KernelKnn` package.

```{r classify, echo=TRUE,cache=TRUE,warning=FALSE,message=F,prompt=FALSE}

# Step 3. Cell classification with KNN.
data = classify.cells(data = data,
                      classes = data$embedded$ccl,
                      k = 11,
                      knn_method = "manhattan",
                      knn_weights_fun = NULL)

# show top ten classified cells
head(data$embedded.predicted)
```

```{r performance, echo=TRUE,cache=TRUE,warning=FALSE,prompt=FALSE}

# Strip from cell name the cell-line
# of origin of each mapped cell
data$embedded.predicted$ccl = sapply(
                            strsplit(x = rownames(data$embedded.predicted),
                                     split = "_",fixed = T)
                            ,function(x) x[1]
                          )

# Now we can compute the classification accuracy
acc = sum(data$embedded.predicted$predicted.class==data$embedded.predicted$ccl)/nrow(data$embedded.predicted)*100
cat("Classification accuracy is",round(acc,2),"%")
```



# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
